---
title       : Transforming Data with `dplyr`
author      : Adam J Sullivan 
job         : Assistant Professor of Biostatistics
work        : Brown University
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js # {highlight.js, prettify, highlight}
hitheme     :  github     # 
widgets     : [mathjax, quiz, bootstrap, interactive] # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: [libraries/nvd3, libraries/leaflet, libraries/dygraphs]}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : publichealthlogo.png
biglogo     : publichealthlogo.png
assets      : {assets: ../../assets}







---  .segue bg:grey

# Piping or Chaining Data

--- .class #id

## Piping or Chaining

- We will discuss a concept that will help us greatly when it comes to working with our data. 
- The usual way to perform multiple operations in one line is by nesting. 

--- .class #id

## Piping or Chaining

To consider an example we will look at the data provided in the gapminder package:

```{r}
library(gapminder)
head(gapminder)
```

--- .class #id

## Nesting vs Chaining

- Let's say that we want to have the GDP per capita and life expectancy Kenya.
- Traditionally speaking we could do this in a nested manner:


```{r, eval=F}
filter(select(gapminder, country, lifeExp, gdpPercap), country=="Kenya")
```

--- .class #id

## Nesting vs Chaining

- It is not easy to see exactly what this code was doing but we can write this in a manner that follows our logic much better. 
- The code below represents how to do this with chaining. 

```{r, eval=F}
gapminder %>%
    select(country, lifeExp, gdpPercap) %>%
    filter(country=="Kenya")
```

--- .class #id

## Breaking Down the Code

- We now have something that is much clearer to read.
- Here is what our chaining command says:
    1. Take the `gapminder` data
    2. Select the variables: `country`, `lifeExp` and `gdpPercap`.
    3. Only keep information from Kenya. 
- The nested code says the same thing but it is hard to see what is going on if you have not been coding for very long.


--- .class #id

## Breaking Down the Code

- The result of this search is below: 


```
# A tibble: 12 x 3
   country lifeExp gdpPercap
    <fctr>   <dbl>     <dbl>
 1   Kenya  42.270  853.5409
 2   Kenya  44.686  944.4383
 3   Kenya  47.949  896.9664
 4   Kenya  50.654 1056.7365
 5   Kenya  53.559 1222.3600
 6   Kenya  56.155 1267.6132
 7   Kenya  58.766 1348.2258
 8   Kenya  59.339 1361.9369
 9   Kenya  59.285 1341.9217
10   Kenya  54.407 1360.4850
11   Kenya  50.992 1287.5147
12   Kenya  54.110 1463.2493
```

--- .class #id

## What is `%>%`

- In the previous code we saw that we used `%>%` in the command you can think of this as saying ***then***.
- For example:

```{r, eval=F}
gapminder %>%
    select(country, lifeExp, gdpPercap) %>%
    filter(country=="Kenya")
```

--- .class #id

## What Does this Mean?

- This translates to:
    - Take Gapminder ***then*** select these columns select(country, lifeExp, gdpPercap) ***then*** filter out so we only keep Kenya


--- .class #id

## Why Chain?

- We still might ask why we would want to do this. 
- Chaining increases readability significantly when there are many commands. 
- With many packages we can replace the need to perform nested arguments. 
- The chaining operator is automatically imported from the [magrittr](https://github.com/smbache/magrittr) package. 

--- .class #id

## User Defined Function


- Let's say that we wish to find the Euclidean distance between two vectors say, `x1` and `x2`. 
- We could use the math formula:

\[\sqrt{\sum(x_1-x_2)^2}\]

--- .class #id

## User Defined Function

- In the nested manner this would be:

```{r results='hide'}
x1 <- 1:5; x2 <- 2:6
sqrt(sum((x1-x2)^2))
```

--- .class #id

## User Defined Function


- However, if we chain this we can see how we would perform this mathematically. 
```{r, eval=FALSE}
# chaining method
(x1-x2)^2 %>% sum() %>% sqrt()
```
- If we did it by hand we would perform elementwise subtraction of `x2` from `x1` ***then*** we would sum those elementwise values ***then*** we would take the square root of the sum. 


--- .class #id

## User Defined Function


```{r}
# chaining method
(x1-x2)^2 %>% sum() %>% sqrt()
```

- Many of us have been performing calculations by this type of method for years, so that chaining really is more natural for us. 


---  .segue bg:grey


# The `dplyr` Package

---  .segue bg:grey

# The `spread()` Function

--- .class #id


## The `spread()` Function

- The first `tidyr` function we will look into is the `spread()` function. 
- With `spread()` it does similar to what you would expect.  
- We have a data frame where some of the rows contain information that is really a variable name. 
- This means the columns are a combination of variable names as well as some data. 


--- .class #id

## The `dplyr` Package

- Now that we have started to tidy up our data we can see that we have a need to transform this data. 
- We may wish to add additional variables. 
- The `dplyr` package allows us to further work with our data. 

--- .class #id


## `dplyr` Functionality

- With `dplyr` we have five basic verbs that we will learn to work with:
  - `filter()`
  - `select()`
  - `arrange()`
  - `mutate()`
  - `summarize()`


  


---  .segue bg:grey


# Filtering 


--- .class #id

## Filtering


- At this point we will consider how we pick the rows of the data that we wish to work with. 
- If you consider many modern data sets, we have so much information that we may not want to bring it all in at once. 
- R brings data into the RAM of your computer. This means you can be limited for what size data you can bring in at once. 
- Very rarely do you need the entire data set. 
- We will focus on how to pick the rows or observations we want now.


--- .class #id

## Enter the `filter()` Function

- The `filter()` function chooses rows that meet a specific criteria. 
- We can do this with Base R functions or with   `dplyr`. 

```{r}
library(dplyr)
```


--- .class #id

## Filtering Example: The Old Way

- Let's say that we want to look at the data just for the country of Kenya in 2002


```{r}
gapminder[gapminder$country=="Kenya" & gapminder$year==2002, ]
```

--- .class #id

## Filtering Example


- Now this is not very difficult to do, what we have is that we are working with `gapminder` and we only want to keep the rows of data there `country=="Kenya` and `year==2002`. 
- However we could use the `filter()` function to do this in a much easier to read format:


--- .class #id

## Filtering Example: Tidyverse Way

- This means in our example we could perform the following:

```
gapminder %>%
    filter(country=="Kenya", year==2002)
```

Finally we could also only do one filtering at a time and chain it:

```
gapminder %>%
    filter(country=="Kenya") %>%
    filter(year==2002)
```


--- .class #id


## Further Filtering

- `filter()` supports the use of multiple conditions where we can use Boolean. 
- For example if we wanted to consider only observations that have a life expectancy between 49 and 60, we could do the following:

```{r, eval=FALSE}
gapminder %>% 
  filter(lifeExp>=49 & lifeExp<60)
```

--- .class #id


## Further Filtering

```{r, echo=FALSE}
gapminder %>% 
  filter(lifeExp>=49 & lifeExp<60)
```

--- .class #id


## Further Filtering

- We can also use the `filter()` function to remove missing data for us. 
- Previously we learned about a class of functions called `is.`*foo*`()` where *foo* represents a data type. 
- We could choose to only use observations that have life expectancy. 
- That means we wish to not have missing data for life expectancy. 


```{r, eval=FALSE}
gapminder %>% filter(!is.na(lifeExp))
```

--- .class #id


## On Your Own: RStudio Practice #5


Using the `filter()` function and chaining: 

- Choose only rows associated with
  - Uganda
  - Rwanda

--- .class #id


## On Your Own: RStudio Practice #5



Your end result should be:

```{R, echo=FALSE}
filter(gapminder, country=="Uganda" | country=="Rwanda")
```

---  .segue bg:grey



# Selecting


--- .class #id

## Selecting

- The next logical step would be to select the columns we want as well. 
- Many times we have so many columns that we are no interested in for a particular analysis. 
- Instead of slowing down your analysis by continuing to run through extra data, we could just select the columns we care about. 

--- .class #id


## Enter the `select()` Function

- The `select()` function chooses columns that we specify. 
- Again we can do this with base functions or with `dplyr`. 
- We feel that as you continue on with your R usage that you will most likely want to go the route of `dplyr` functions instead.

--- .class #id

## Select Example: Old Way

- Let's say that we want to look at the gapminder data but we are really only interested in the country, life expectancy and year. 
- This seems reasonable if we are a customer and wanted to only know these pieces of information. We could do this with indexing : 

```{r, eval=FALSE}
gapminder[, c("country", "lifeExp", "year")]
```


--- .class #id

## Selecting Example: Tidyverse Way 

We could then do the following

```{r,eval= FALSE}
gapminder %>%
  select(country, lifeExp, year)
```


--- .class #id

## Selecting Example: Tidyverse Way 



```{r,echo= FALSE}
gapminder %>%
  select(country, lifeExp, year)
```


--- .class #id

## Removing Columns


- We may wish to pick certain columns that we wish to have but we also may want to remove certain columns. 
- It is quite common to de-identify a dataset before actually distributing it to a research team. 
- The `select()` function will also remove columns. 



--- .class #id

## Removing Columns

- Lets say that we wished to remove the `gdpPercap` and `pop` of the countries:

```{R, eval=FALSE}
gapminder %>%
  select(-gdpPercap,-pop)
```

--- .class #id

## Removing Columns

We also could use a vector for this:

```{R, eval=F}

cols <- c("gdpPercap", "pop")
gapminder %>%
  select(-one_of(cols))
```

--- .class #id

## Removing Columns

- We can also remove columns that contain a certain phrase in the name. 
- If we were interested in removing any columns that had to do with time we could search for the phrase "co" in the data and remove them:

```{r, eval= FALSE}
gapminder %>%
  select(-matches("Per"))
```

--- .class #id

## Removing Columns


```{r, echo= FALSE}
gapminder %>%
  select(-matches("Per"))
```

--- .class #id

## Unique Observations

- Many times we have a lot of repeats in our data. 
- If we just would like to have an account of all things included then we can use the `unique()` command. 
- Lets assume that we wish to know which countries are in the data. 
- We do not want to have every country listed over and over again so we ask for unique values:

```{r, eval=F}
gapminder %>% 
  select(country) %>% 
  unique()
```


--- .class #id

## On Your Own: RStudio Practice # 6

1. Select all but the `pop` column.
2. Remove the year and lifeExp from this new dataframe.
3. Select values which contain "p" in them. 
4. Chain these together so that you run a command and it does all of these things. 



--- .class #id

## On Your Own: RStudio Practice # 6


Your answer should look like:

```{R, echo=FALSE}
gapminder %>% select(-pop)  %>% select(-lifeExp,-year) %>% select(contains("p"))
```


---  .segue bg:grey

# Arranging the Data

--- .class #id

## Arranging the Data

- We also have need to make sure the data is ordered in a certain manner. This can be easily done in R with the `arrange()` function. 
- Again we can do this in base R but this is not always a clear path. 

--- .class #id

## Arranging the Data Example: Old Way

- Let's say that we wish to look countries, year and life expectancy. 
- Thensay further we want to sort it by Life Expactancy. 
- In base R we would have to run the following command:

```{r results='hide'}
library(gapminder)
library(tidyverse)
gapminder[order(gapminder$lifeExp), c("country", "year", "lifeExp")]
```



--- .class #id

## Arranging the Data Example: Tidyverse Way
```{r, eval=FALSE}

gapminder %>%
    select(country,year,  lifeExp) %>%
    arrange(lifeExp)
```

--- .class #id

## Arranging the Data Example: Tidyverse Way

```{r, echo=FALSE}


gapminder %>%
    select(country,year,  lifeExp) %>%
    arrange(lifeExp)
```

--- .class #id

## Arranging the Data Example: Tidyverse Way

- With `arrange()` we first use `select()` to pick the only columns that we want and then we arrange by the `lifeExp`. 
- If we had wished to order them in a descending manner we could have simply used the `desc()` function:

```{r, eval=FALSE }
gapminder %>%
    select(country,year,  lifeExp) %>%
    arrange(desc(lifeExp))
```

--- .class #id


## More Complex Arrange

- Lets consider that we wish to look at the top 3 Life Expectancies for each year.
- Then we wish to order them from largest to smallest Life Expectancy. 
- We then need to do the following:
    1. Group by year.
    2. Pick the top 3 life expectancy
    3. order them largest to smallest
    



--- .class #id

## More Complex Arrange Continued


```{R, eval=FALSE}
gapminder %>% 
  group_by(year) %>%  
  top_n(3, lifeExp) %>% 
  arrange(desc(lifeExp))
```

- Where
    - `group_by()` is a way to group data. This way we perform operations on a group. So top 3 life expectancy are grouped by year. 
    - `top_n()`takes a tibble and returns a specific number of rows based on a chosen value. 


--- .class #id

## More Complex Arrange Continued


```{R, echo=FALSE}
gapminder %>% 
  group_by(year) %>%  
  top_n(3, lifeExp) %>% 
  arrange(desc(lifeExp))
```



--- .class #id

## On Your Own: RStudio Practice #7 

- Perform the following operations:
  - Group by year. 
  - use `sample_n()` to pick 1 observation per year 
  - Arrange by longest to smallest life expectancy. 



--- .class #id





## On Your Own: RStudio Practice


Your answer ***may*** look like:

```{r, echo=FALSE}
gapminder %>%
  group_by(year) %>%
  sample_n(1) %>%
  arrange(desc(lifeExp))
```

---  .segue bg:grey


# Summarizing Data


--- .class #id

## Summarizing Data

- As you have seen in your own work, being able to summarize information is crucial. 
- We need to be able to take out data and summarize it as well. 
- We will consider doing this using the `summarise()` function. 

--- .class #id

## Summarizing Data: Old Way

- Like in the rest of these lessons, let's consider what happens when we try to to do this in base R. We will:
  1. Create a table grouped by `country`.
  2. Summarize each group by taking mean of `lifeExp`.

```{r results='hide'}
head(with(gapminder, tapply(lifeExp, country, mean, na.rm=TRUE)))
head(aggregate(lifeExp ~ country, gapminder, mean))
```

--- .class #id


## Enter `summarise()` Function

- The `summarise()` function is:

```
summarise(.data, ...)
```

- where
  - `.data` is the tibble of interest.
  - `...` is a list of name paired summary functions
    - Such as:
      - `mean()`
      - `median`
      - `var()`
      - `sd()`
      - `min()`
      - `max()
      - ...



--- .class #id

## Summarizing Data Example: Tidyverse Way 

- Consider the logic here:
  1. Group data by country
  2. Find the average life Expectancy of the groups and call it `avg_lifeExp`.
- This is much easier to understand than the  Base R code. 

```{r, eval=FALSE}
gapminder %>%
    group_by(country) %>%
    summarise(avg_lifeExp = mean(lifeExp, na.rm=TRUE))
```

--- .class #id


## Summarizing Data Example: Tidyverse Way 


```{r, echo=FALSE}
gapminder %>%
    group_by(country) %>%
    summarise(avg_lifeExp = mean(lifeExp, na.rm=TRUE))
```

--- .class #id


## Another Example

- Lets say that we would like to have more than just the averages but we wish to have the minimum and the maximum life expectancies by country:

```{r, eval=FALSE}

gapminder %>%
    group_by(country) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), lifeExp)
```


--- .class #id


## Another Example



```{r, echo=FALSE}
gapminder %>%
    group_by(country) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), lifeExp)
```

--- .class #id

## On Your Own: RStudio Practice #8 

- The following is a new function:
  - Helper function `n()` counts the number of rows in a group
- Then for each year and continent:
  - count total lifeExp
  - Sort in descending order. 

--- .class #id

## On Your Own: RStudio Practice # 8 

- Your answer should look like:

```{r, eval=FALSE}

gapminder %>%
    group_by(continent,year) %>%
    summarise(lifeExp_count = n()) %>%
    arrange(desc(lifeExp_count))
```

--- .class #id

## On Your Own: RStudio Practice # 8 

- We could also have used what is called the  `tally()` function:

```{r}
gapminder %>%
    group_by(country, year) %>%
    tally(sort = TRUE)
```

---  .segue bg:grey


# Adding New Variables


--- .class #id

## Adding New Variables

- There is usually no way around needing a new variable in your data. 
- For example, most medical studies have height and weight in them, however many times what a researcher is interested in using is Body Mass Index (BMI). 
- We would need to add BMI in. 


--- .class #id

## Adding New Variables

- Using the `tidyverse` we can add new variables in multiple ways
  - `mutate()`
  - `transmute()`



--- .class #id


## Differences Between `mutate()` and `transmute()`

- There is only one major difference between `mutate()` and `transmutate` and that is what it keeps in your data. 
  - `mutate()`
      - creates a new variable
      - It keeps all existing variables
  - `transmute()`
      - creates a new variable.
      - It only keeps the new variables


--- .class #id

## Example

- Let's say we wish to have a variable called gdp We want to basically do:

\[\text{gdp} = gdpPercap\times pop\]


- We can first do this with `mutate()`:

```{r, eval=FALSE}
gapminder %>% 
  select(country, gdpPercap, pop) %>%
  mutate(gdp = gdpPercap*pop)
```

--- .class #id

## Example: Mutate

```{r, echo=FALSE}
gapminder %>% 
  select(country, gdpPercap, pop) %>%
  mutate(gdp = gdpPercap*pop)
```

--- .class #id

## Example: Transmute


```{r, eval=FALSE}
gapminder %>% 
  select(country, gdpPercap, pop) %>%
  transmute(gdp = gdpPercap*pop)
```


--- .class #id

## Example: transmute


```{r, echo=FALSE}
gapminder %>% 
  select(country, gdpPercap, pop) %>%
  transmute(gdp = gdpPercap*pop)
```

---  .segue bg:grey



# Further Summaries

--- .class #id

## Further Summaries

- We have so far discussed how one could find the basic number summaries:
  - mean
  - median
  - standard deviation
  - variance
  - minimum 
  - maximum
- However there are many more operations that you may wish to do for summarizing data. 
- In fact many of the following examples are excellent choices for working with categorical data which does not always make sense to do the above summaries for. 


--- .class #id

## Further Summaries

- We will consider:
  1. Grouping and Counting
  2. Grouping, Counting and Sorting
  3. Other Groupings
  4. Counting Groups


--- .class #id



## Grouping and Counting


- We have seen the functions `tally()` and `count()`. 
- Both of these can be used for grouping and counting. 
- They also are very concise in how they are called. 


--- .class #id



## Grouping and Counting


- For example if we wished to know how many countrries there were by year, we would use `tally()` in this manner: 

```{r, eval=FALSE }
gapminder %>%
  group_by(year) %>% 
  tally()
```

--- .class #id



## Grouping and Counting

- Where as we could do the same thing with `count()`

```{r, eval=FALSE}
gapminder %>% 
  count(year)
```

*Notice: `count()` allowed for month to be called inside of it, removing the need for the `group_by()` function. 


--- .class #id



## Grouping, counting and sorting.

- Both `tally()` and `count()` have an argument called `sort()`. 
- This allows you to go one step further and group by, count and sort at the same time. 
- For `tally()` this would be:

```{r, eval=FALSE}
gapminder %>% group_by(year) %>% tally(sort=TRUE)
```

--- .class #id

## Grouping, counting and sorting: `tally()`

```{r, echo=FALSE}
gapminder %>% group_by(year) %>% tally(sort=TRUE)
```




--- .class #id

## Grouping with other functions

- We can also sum over other values rather than just counting the rows like the above examples. 
- For example let us say we were interested in know the total GDP for countries in a given continent and year.
- We could do this with the `summarise()` function, `tally()` function or the `count()` function:

```{r, eval=FALSE}
gapminder %>% 
  group_by(continent, year) %>% 
  summarise(total_gdp = sum(gdp))
```


--- .class #id

## Grouping with other functions


```{r, echo=FALSE}
gapminder %>% 
  group_by(continent, year) %>% 
  summarise(total_gdp = sum(gdp))
```


--- .class #id

## What happened?

- Where did gdp go?
- We made it before but we did not save it. 

--- .class #id

## Saving our work

```{r}
gapminder <- gapminder %>%
    mutate(gdp=gdpPercap*pop)
```

--- .class #id

## Try again:


```{r, echo=FALSE}
gapminder %>% 
  group_by(continent) %>%
  summarise(total_gdp = sum(gdp))
```


--- .class #id

## Grouping with other functions

- For  `tally()` we could do:

```{r ,eval=FALSE}
gapminder %>% 
  group_by(continent) %>% 
  tally(wt = gdp)
```

*Note: in `tally()` the `wt` stands for weight and allows you to weight the sum based on the gdp*. 

--- .class #id

## Grouping with other functions

- With the `count()` function we also use `wt`:


```{r}
gapminder %>% count(continent, wt = gdp)
```

--- .class #id

## Counting Groups


- We may want to know how large our groups are. To do this we can use the following functions:
  - `group_size()` is a function that returns counts of group. 
  - `n_groups()` returns the number of groups

--- .class #id

## Counting Groups

- So if wanted to count the number of countries by continent, we could group by continent and find the groups size using `group_size()`:

```{r, eval=FALSE}
gapminder %>% 
  group_by(continent) %>% 
  group_size()
```


--- .class #id

## Counting Groups



```{r, echo=FALSE}
gapminder %>% 
  group_by(continent) %>% 
  group_size()
```


--- .class #id

## Counting Groups


- If we just wished to know how many years were represented in our data we could use the `n_groups()` function:


```{r, eval=FALSE}
gapminder %>% 
  group_by(year) %>% 
  n_groups()
```



--- .class #id

## Counting Groups

```{r, echo=FALSE}
gapminder %>% 
  group_by(year) %>% 
  n_groups()
```

