---
title: "02 Summarizing Data with `dplyr` Practice"
output: html_notebook
---

## Setup

The first chunk in an R Notebook is usually titled "setup," and by convention includes the R packages you want to load. Remember, in order to use an R package you have to run some `library()` code every session. Execute these lines of code to load the packages. 

```{r setup}
library(ggplot2)
library(gapminder)
```




## On Your Own: RStudio Practice #5


Using the `filter()` function and chaining: 

- Choose only rows associated with
  - Uganda
  - Rwanda
  
  
```{r}

gapminder %>% 
  filter()
```



## On Your Own: RStudio Practice # 6

1. Select all but the `pop` column.
2. Remove the year and lifeExp from this new dataframe.
3. Select values which contain "p" in them. 
4. Chain these together so that you run a command and it does all of these things. 



```{R, echo=FALSE}
gapminder 
```



## On Your Own: RStudio Practice #7 

- Perform the following operations:
  - Group by year. 
  - use `sample_n()` to pick 1 observation per year 
  - Arrange by longest to smallest life expectancy. 





```{r, eval=FALSE}
gapminder %>%
  
```



## On Your Own: RStudio Practice #8 

- The following is a new function:
  - Helper function `n()` counts the number of rows in a group
- Then for each year and continent:
  - count total lifeExp
  - Sort in descending order. 



```{r, echo=FALSE}

gapminder %>%
   
```
